---
title: "CD8_APLcombined_analysis_2017"
author: "Arianne Richard"
date: "March 2018"
output: html_document
---

This file walks through all of the analyses combining the single-cell transcriptomic data from both runs to gain more power for particular analyses.

First we load the data, remove uninformative genes, and normalize the data together.


```{r}
library(scater)
library(scran)
library(SingleCellExperiment)
library(RColorBrewer)
library(gplots)
library(dpt)
source('dpt_plots.R')


sce1 <- readRDS('../data/sce_run1.rds')
sce2 <- readRDS('../data/sce_run2.rds')

## filter to remove non-expressed genes -- require detection in at least 5% of cells in each batch
drop_filt <- rowSums(counts(sce1) > 0) >= (ncol(sce1) * 0.05)
sce1 <- sce1[drop_filt,]
mean_filt <- rowMeans(counts(sce1)) >= 1
sce1 <- sce1[mean_filt,]

drop_filt <- rowSums(counts(sce2) > 0) >= (ncol(sce2) * 0.05)
sce2 <- sce2[drop_filt,]
mean_filt <- rowMeans(counts(sce2)) >= 1
sce2 <- sce2[mean_filt,]

## add a phenoData column that reflects both batch and plate
colData(sce1)$batch <- paste0('batch1 ', colData(sce1)$Sorting.Plate)
colData(sce2)$batch <- paste0('batch2 ', colData(sce2)$Sorting.Plate)

## make them all have the same phenoData columns
colData(sce1) <- colData(sce1)[,colnames(colData(sce1)) %in% colnames(colData(sce2))]
colData(sce2) <- colData(sce2)[,colnames(colData(sce2)) %in% colnames(colData(sce1))]

## make them all have the same featureData columns
rowData(sce1) <- rowData(sce1)[colnames(rowData(sce1)) %in% colnames(rowData(sce2)),]
rowData(sce2) <- rowData(sce2)[colnames(rowData(sce2)) %in% colnames(rowData(sce1)),]

## make sure all genes same
sce1 <- sce1[rownames(sce1) %in% rownames(sce2),]
sce2 <- sce2[rownames(sce2) %in% rownames(sce1),]
table(rownames(sce1)==rownames(sce2))

## mergeSCESets
sce <- cbind(sce1, sce2)

## calculate spike factors for the data
sce <- computeSpikeFactors(sce, general.use=TRUE)

## normalise the data
sce <- normalize(sce, exprs_values='counts', return_norm_as_exprs=TRUE, log_exprs_offset=1)

## let's look at the spike factors
plot(log10(sizeFactors(sce)), log10(sce$mapped), xlab='ERCC factor', ylab='total counts',
     col=c('blue', 'darkgreen','red', 'orange', 'purple', 'magenta')[factor(sce$batch)])
legend('bottomright', legend=levels(factor(sce$batch)), 
       fill=c('blue', 'darkgreen','red', 'orange', 'purple', 'magenta'), bty='n')

```

The second batch seems to have fewer reads but a similar ratio of cellular to ERCC reads.   Now let's correct for any batch effects.

```{r}
## let's look at the batch effect
pcs <- prcomp(t(as.matrix(exprs(sce[!isSpike(sce),]))), scale=TRUE)
par(mgp=c(2,1,0), bty='l')
plot(pcs$x[,1], pcs$x[,2], xlab='PC1', ylab='PC2', 
     col=c('blue', 'darkgreen','red', 'orange', 'purple', 'magenta')[factor(sce$batch)], pch=16, cex.lab=1.2)
legend('topleft', legend=levels(factor(sce$batch)), 
       fill=c('blue', 'darkgreen','red', 'orange', 'purple', 'magenta'), ncol=2, bty='n')

## then do the batch correction
library(sva)
mod <- model.matrix(~Condition, data=colData(sce))
combat_sce <- ComBat(dat=exprs(sce), batch=colData(sce)$batch, mod=mod, par.prior=TRUE, prior.plots=FALSE)
exprs(sce) <- combat_sce

## and check it again
pcs <- prcomp(t(as.matrix(exprs(sce[!isSpike(sce),]))), scale=TRUE)
par(mgp=c(2,1,0), bty='l')
plot(pcs$x[,1], pcs$x[,2], xlab='PC1', ylab='PC2', 
     col=c('blue', 'darkgreen','red', 'orange', 'purple', 'magenta')[factor(sce$batch)], pch=16, cex.lab=1.2)
legend('bottomleft', legend=levels(factor(sce$batch)), 
       fill=c('blue', 'darkgreen','red', 'orange', 'purple', 'magenta'), ncol=2, bty='n')

## and finally look by condition or protein expression

plotPCs <- function(x, cols, legend='', legend_cols){
  par(mfrow=c(2,2), mgp=c(2,0.5,0), mar=c(3,3.5,0.2,0.2), bty='l')
  plot(x[,1], x[,2], xlab='', ylab='PC2', pch=16, col=cols, cex.lab=1.5)
  plot.new()
  if(legend[1] !=''){
    legend('center', legend=legend, 
         fill=legend_cols, 
         bty='n', ncol=2)
  }
  plot(x[,1], x[,3], xlab='PC1', ylab='PC3', pch=16, col=cols, cex.lab=1.5)
  plot(x[,2], x[,3], xlab='PC2', ylab='', pch=16, col=cols, cex.lab=1.5)
}


cols <- as.character(colData(sce)$Condition)
cols[cols %in% 'N4_6h'] <- 'red'
cols[cols %in% 'N4_3h'] <- 'orange'
cols[cols %in% 'N4_1h'] <- 'gold'
cols[cols %in% 'unstimulated'] <- 'grey'
cols[cols %in% 'T4_6h'] <- 'green3'
cols[cols %in% 'G4_6h'] <- 'blue'
cols[cols %in% 'NP68_6h'] <- 'grey40'

plotPCs(pcs$x, cols, legend=c('0h', '1h', '3h', '6h', 'T4 6h', 'G4 6h', 'NP68 6h'), 
        legend_cols=c('grey', 'gold', 'orange', 'red', 'green3', 'blue', 'grey40'))

pdf('../data/PCs_combrun_spikenorm_filt05.pdf', height=8, width=8)
plotPCs(pcs$x, cols, legend=c('0h', '1h', '3h', '6h', 'T4 6h', 'G4 6h', 'NP68 6h'), 
        legend_cols=c('grey', 'gold', 'orange', 'red', 'green3', 'blue', 'grey40'))
dev.off()

## Now we plot PCs coloured by protein expression, except that we must plot only one sequencing run to avoid date-specific differences in flow protein measurements.

library(viridis)
colCD69 <- viridis(100)[as.numeric(cut(as.numeric(as.character(colData(sce[,sce$Date %in% '2016-10-26'])$Characteristics.CD69.measurement..log10..)),breaks = 100))]
colCD25 <- viridis(100)[as.numeric(cut(as.numeric(as.character(colData(sce[,sce$Date %in% '2016-10-26'])$Characteristics.CD25.measurement..log10..)),breaks = 100))]
colCD62L <- viridis(100)[as.numeric(cut(as.numeric(as.character(colData(sce[,sce$Date %in% '2016-10-26'])$Characteristics.CD62L.measurement..log10..)),breaks = 100))]
colCD44 <- viridis(100)[as.numeric(cut(as.numeric(as.character(colData(sce[,sce$Date %in% '2016-10-26'])$Characteristics.CD44.measurement..log10..)),breaks = 100))]

plotPCs(pcs$x[sce$Date %in% '2016-10-26',], cols=colCD69)
plotPCs(pcs$x[sce$Date %in% '2016-10-26',], cols=colCD25)
plotPCs(pcs$x[sce$Date %in% '2016-10-26',], cols=colCD62L)
plotPCs(pcs$x[sce$Date %in% '2016-10-26',], cols=colCD44)


```

Let's now look at diffusion pseudotime. We learned from the time-course data that there is a TCR-independent burst of TF activity at 1 hour, so we'll exclude that time-point.

```{r}
library(destiny)

## first get rid of spike-ins
sce <- sce[!isSpike(sce),]
sceAT <- sce[,sce$Condition %in% c('N4_6h', 'T4_6h', 'G4_6h', 'NP68_6h', 'unstimulated', 'N4_3h')]
resAT <- quick_dpt(sceAT, conditions=c('N4_6h', 'T4_6h', 'G4_6h', 'NP68_6h', 'N4_3h', 'unstimulated'), col=c('red', 'green3', 'blue', 'grey40', 'orange', 'grey'), base_condition='unstimulated', stim_condition='N4_6h', trajectory_name='transcriptomic activation status')

resAT[[1]]

resAT[[2]]

pdf('../data/DPT_combrunAT_spikenorm_filt05_AT.pdf', height=2.5, width=4)
resAT[[1]]
dev.off()

pdf('../data/DPT_combrunAT_spikenorm_filt05_AT_distribution.pdf', height=2.5, width=4)
resAT[[2]]
dev.off()

```

It looks like most cells are following the same trajectory regardless of stimulus.  Let's formally test for this by controlling for activation status while performing a differential expression analysis by stimulus. To do this robustly, we should have an independent marker of activation (not progression along the pseudotime trajectory because this would be overift). Both CD69 and CD25 would be robust markers of classical cellular activation and could serve as a surrogate for activation status when we do differential expression analysis accounting for activation. Let's look at which protein expression best agrees with transcriptomic activation status as defined by pseudotime progression.

```{r}

plot_DPTvprotein <- function(sceset, sort_date, DPTres, protein, trajectory_name='pseudotime'){
  coln <- paste0('Characteristics.', protein, '.measurement..log10..')
  cort <- cor.test(as.numeric(as.character(colData(sceset)[sceset$Date %in% sort_date, coln])), 
       DPTres$pt[sceset$Date %in% sort_date,'DPT'])
  plot(as.numeric(as.character(colData(sceset)[sceset$Date %in% sort_date, coln])), 
       DPTres$pt[sceset$Date %in% sort_date,'DPT'], 
       pch=20, 
       xlab=bquote('log'[10] ~ .(protein) ~ 'expression'), 
       ylab=trajectory_name, 
       cex=0.8, las=1)
  text(x=max(as.numeric(as.character(colData(sceset)[sceset$Date %in% sort_date, coln])))-0.25,
       y=min(DPTres$pt[sceset$Date %in% sort_date,'DPT'])+0.05, 
       labels=bquote(italic('r ') ~ .(round(cort$estimate, 2))))
}

par(mfrow=c(2,2), bty='l')
plot_DPTvprotein(sceAT, '2016-10-26', resAT, 'CD69', trajectory_name='transcriptional activation status')
plot_DPTvprotein(sceAT, '2016-07-27', resAT, 'CD69', trajectory_name='transcriptional activation status')
plot_DPTvprotein(sceAT, '2016-10-26', resAT, 'CD25', trajectory_name='transcriptional activation status')
plot_DPTvprotein(sceAT, '2016-07-27', resAT, 'CD25', trajectory_name='transcriptional activation status')

pdf('../data/DPT_combrunAT_spikenorm_filt05_DPTvCD25_2.pdf', height=4.5, width=4)
par(mgp=c(2.5,0.5,0))
plot_DPTvprotein(sceAT, '2016-10-26', resAT, 'CD25', trajectory_name='transcriptional activation status')
dev.off()

pdf('../data/DPT_combrunAT_spikenorm_filt05_DPTvCD25_1.pdf', height=4.5, width=4)
par(mgp=c(2.5,0.5,0))
plot_DPTvprotein(sceAT, '2016-07-27', resAT, 'CD25', trajectory_name='transcriptional activation status')
dev.off()

pdf('../data/DPT_combrunAT_spikenorm_filt05_DPTvCD69_2.pdf', height=4.5, width=4)
par(mgp=c(2.5,0.5,0))
plot_DPTvprotein(sceAT, '2016-10-26', resAT, 'CD69', trajectory_name='transcriptional activation status')
dev.off()

pdf('../data/DPT_combrunAT_spikenorm_filt05_DPTvCD69_1.pdf', height=4.5, width=4)
par(mgp=c(2.5,0.5,0))
plot_DPTvprotein(sceAT, '2016-07-27', resAT, 'CD69', trajectory_name='transcriptional activation status')
dev.off()

```

CD69 is very well associated with pseudotime and is generally accepted as a marker of activation. Its only draw back is that it doesn't have great overlap between G4 and N4.  To avoid assuming linearity, we'll use a spline fit to protein expression - hopefully this will help us avoid spurious results, when controlling for CD69 expression in DE analysis between peptides. We'll also include an interaction term with the date of CD69 protein expression measurements because these FACS values are on a relative scale and are not directly comparable between experiments.

```{r}
AT_mat <- data.frame(Cond=colData(sceAT)$Condition, Batch=colData(sceAT)$batch)
AT_mat$Cond <- unlist(lapply(strsplit(as.character(AT_mat$Cond), split='_'), function(x){x[1]}))
AT_mat$Cond <- factor(AT_mat$Cond, levels=c('N4', 'T4', 'G4', 'NP68', 'unstimulated'))
AT_mat$CD69 <- as.numeric(as.character(colData(sceAT)$Characteristics.CD69.measurement..log10..))
AT_mat$Date <- factor(sceAT$Date)

library(splines)
CD69sp2 <- ns(AT_mat$CD69[sceAT$Date %in% '2016-10-26'], df=4)
rownames(CD69sp2) <- rownames(colData(sceAT))[sceAT$Date %in% '2016-10-26']
CD69sp1 <- ns(AT_mat$CD69[sceAT$Date %in% '2016-07-27'], df=4)
rownames(CD69sp1) <- rownames(colData(sceAT))[sceAT$Date %in% '2016-07-27']
CD69spdf <- rbind(CD69sp2, CD69sp1)
CD69spdf <- CD69spdf[rownames(colData(sceAT)),]

mod <- model.matrix(~1+AT_mat$Batch + AT_mat$Date:CD69spdf + AT_mat$Cond) 

library(edgeR)
z_AT <- convertTo(sceAT, type="edgeR")
y_s <- estimateDisp(z_AT, design=mod)
fit <- glmFit(z_AT, design=mod, dispersion=y_s$tagwise.dispersion, prior.count=1)
resT4 <- glmLRT(fit, coef='AT_mat$CondT4')
bestT4 <- data.frame(topTags(resT4, n=nrow(z_AT)))
fdat <- rowData(sceAT)
rownames(fdat) <- rownames(sceAT)
bestT4$symbol <- as.character(fdat[rownames(bestT4),'symbol'])
bestT4[bestT4$FDR < 0.05,]
write.table(bestT4, file='../data/CombrunAT_spikenorm_filt05_T4vN4_controlCD69sp2.txt', sep='\t', quote=F)

resG4 <- glmLRT(fit, coef='AT_mat$CondG4')
bestG4 <- data.frame(topTags(resG4, n=nrow(z_AT)))
bestG4$symbol <- as.character(fdat[rownames(bestG4),'symbol'])
bestG4[bestG4$FDR < 0.05,]
write.table(bestG4, file='../data/CombrunAT_spikenorm_filt05_G4vN4_controlCD69sp2.txt', sep='\t', quote=F)

```

Let's make some volcano plots to see these results.

```{r}
## volcano plots of G4 v N4 and T4 v N4 with CD69 covariate 

par(bty='l', mgp=c(2.5,1,0))
plot(bestG4$logFC, -log10(bestG4$PValue), pch=20, cex=0.7, 
     xlab=expression(paste("log"[2], "(fold-change)")),
     ylab=expression(paste("-log"[10], "(p value)")),
     cex.lab=1.6, cex.axis=1.6)
points(bestG4$logFC[which(bestG4$FDR < 0.05)], -log10(bestG4$PValue[which(bestG4$FDR < 0.05)]), 
     pch=20, cex=0.6, col='magenta') 

par(bty='l', mgp=c(2.5,1,0))
plot(bestT4$logFC, -log10(bestT4$PValue), pch=20, cex=0.7, 
     xlab=expression(paste("log"[2], "(fold-change)")),
     ylab=expression(paste("-log"[10], "(p value)")), 
     cex.lab=1.6, cex.axis=1.6)
points(bestT4$logFC[which(bestT4$FDR < 0.05)], -log10(bestT4$PValue[which(bestT4$FDR < 0.05)]), 
       pch=20, cex=0.6, col='magenta') 

pdf('../data/DE_combrunAT_spikenorm_filt05_G4vN4_controlCD69sp2_volcano.pdf', height=5, width=5)
par(bty='l', mgp=c(2.5,1,0))
plot(bestG4$logFC, -log10(bestG4$PValue), pch=20, cex=0.7, 
     xlab=expression(paste("log"[2], "(fold-change)")),
     ylab=expression(paste("-log"[10], "(p value)")),
     cex.lab=1.6, cex.axis=1.6)
points(bestG4$logFC[which(bestG4$FDR < 0.05)], -log10(bestG4$PValue[which(bestG4$FDR < 0.05)]), 
     pch=20, cex=0.6, col='magenta') 
dev.off()

pdf('../data/DE_combrunAT_spikenorm_filt05_T4vN4_controlCD69sp2_volcano.pdf', height=5, width=5)
par(bty='l', mgp=c(2.5,1,0))
plot(bestT4$logFC, -log10(bestT4$PValue), pch=20, cex=0.7, 
     xlab=expression(paste("log"[2], "(fold-change)")),
     ylab=expression(paste("-log"[10], "(p value)")), 
     cex.lab=1.6, cex.axis=1.6)
points(bestT4$logFC[which(bestT4$FDR < 0.05)], -log10(bestT4$PValue[which(bestT4$FDR < 0.05)]), 
       pch=20, cex=0.6, col='magenta') 
dev.off()
```

How similar are the datasets?  Let's intersect the significant lists.  This certainly suffers from a thresholding problem, but we aren't making any claims about specificity.

```{r}
shared <- rownames(bestT4[bestT4$FDR<0.05,])[rownames(bestT4[bestT4$FDR<0.05,]) 
                                             %in% rownames(bestG4[bestG4$FDR<0.05,])]
write.table(bestT4[shared,'symbol'], '../data/CombrunAT_spikenorm_filt05_bothvN4_control2CD69sp2_significant.txt', sep='\t', quote=F)

bestT4[shared,]

par(bty='l', mgp=c(2,0.5,0))
plot(bestT4$logFC, bestG4[rownames(bestT4),'logFC'], pch=20,
     xlab=expression(paste("log"[2], " fold change T4vN4")), 
     ylab=expression(paste("log"[2], " fold change G4vN4")), 
     cex=0.6, col='lightgrey', cex.axis=0.8, cex.lab=0.8, las=1)
points(bestT4[bestT4$FDR<0.05,'logFC'], bestG4[rownames(bestT4[bestT4$FDR<0.05,]),'logFC'], 
       col='green3', pch=20)
points(bestT4[rownames(bestG4[bestG4$FDR<0.05,]),'logFC'], bestG4[bestG4$FDR<0.05,'logFC'], 
       col='blue', pch=20)
points(bestT4[shared,'logFC'], bestG4[shared,'logFC'], col='magenta', pch=20)
text(bestT4[shared[c(1,2,3,7,8)],'logFC'], bestG4[shared[c(1,2,3,7,8)],'logFC'], 
     labels=bestT4[shared[c(1,2,3,7,8)],'symbol'], cex= 0.7, adj=c(-0.1,0.5), col='black', font=2)
text(bestT4[shared[c(5)],'logFC'], bestG4[shared[c(5)],'logFC'], 
     labels=bestT4[shared[c(5)],'symbol'], cex= 0.7, adj=c(0,1.2), col='black', font=2)
text(bestT4[shared[c(4)],'logFC'], bestG4[shared[c(4)],'logFC'], 
     labels=bestT4[shared[c(4)],'symbol'], cex= 0.7, adj=c(1,-0.2), col='black', font=2)
text(bestT4[shared[c(6)],'logFC'], bestG4[shared[c(6)],'logFC'], 
     labels=bestT4[shared[c(6)],'symbol'], cex= 0.7, adj=c(1,1.2), col='black', font=2)
legend('bottomright', legend=c('T4 v N4', 'G4 v N4', 'both'), 
       fill=c('green3', 'blue', 'magenta'), bty='n', cex=0.8)

pdf('../data/DE_combrunAT_spikenorm_filt05_G4T4vN4_controlCD69sp2_logFCs.pdf', height=4.5, width=4.2)
par(bty='l', mgp=c(2,0.5,0), mar=c(5.1, 4.1, 2.1, 2.1))
plot(bestT4$logFC, bestG4[rownames(bestT4),'logFC'], pch=20,
     xlab=expression(paste("log"[2], " fold change T4vN4")), 
     ylab=expression(paste("log"[2], " fold change G4vN4")), 
     cex=0.6, col='lightgrey', cex.axis=0.8, cex.lab=0.8, las=1, xlim=c(-3, 4))
points(bestT4[bestT4$FDR<0.05,'logFC'], bestG4[rownames(bestT4[bestT4$FDR<0.05,]),'logFC'], 
       col='green3', pch=20)
points(bestT4[rownames(bestG4[bestG4$FDR<0.05,]),'logFC'], bestG4[bestG4$FDR<0.05,'logFC'], 
       col='blue', pch=20)
points(bestT4[shared,'logFC'], bestG4[shared,'logFC'], col='magenta', pch=20)
text(bestT4[shared[c(1,2,3,7,8)],'logFC'], bestG4[shared[c(1,2,3,7,8)],'logFC'], 
     labels=bestT4[shared[c(1,2,3,7,8)],'symbol'], cex= 0.7, adj=c(-0.1,0.5), col='black', font=2)
text(bestT4[shared[c(5)],'logFC'], bestG4[shared[c(5)],'logFC'], 
     labels=bestT4[shared[c(5)],'symbol'], cex= 0.7, adj=c(0,1.2), col='black', font=2)
text(bestT4[shared[c(4)],'logFC'], bestG4[shared[c(4)],'logFC'], 
     labels=bestT4[shared[c(4)],'symbol'], cex= 0.7, adj=c(1,-0.2), col='black', font=2)
text(bestT4[shared[c(6)],'logFC'], bestG4[shared[c(6)],'logFC'], 
     labels=bestT4[shared[c(6)],'symbol'], cex= 0.7, adj=c(1,1.2), col='black', font=2)
legend('bottomright', legend=c('T4 v N4', 'G4 v N4', 'both'), 
       fill=c('green3', 'blue', 'magenta'), bty='n', cex=0.8)
dev.off()



```

Approximately how many genes are differentially expressed between the peptide conditions? In other words, what is pi-0, or the estimate of how often the null hypothesis is true?

```{r}
# library(qvalue)
# p0T4 <- pi0est(p=bestT4$PValue, pi0.method = 'bootstrap')
# p0T4
# p0G4 <- pi0est(p=bestG4$PValue, pi0.method = 'bootstrap')
# p0G4

ps_list <- data.frame(peptides=c(rep('T4', nrow(bestT4)), rep('G4', nrow(bestG4))), ps=c(bestT4$PValue, bestG4$PValue))
ps_list$adj <- p.adjust(ps_list$ps, method='fdr')
ps_list$sig <- ps_list$adj < 0.05
tab <- table(ps_list$sig, ps_list$peptides)
G4sig <- tab['FALSE', 'G4']/sum(tab[,'G4'])
T4sig <- tab['FALSE', 'T4']/sum(tab[,'T4'])

print(paste0('Percent not significantly differentially expressed T4 v N4 is ', T4sig))
print(paste0('Percent not significantly differentially expressed G4 v N4 is ', G4sig))

tab
G4sig <- tab['TRUE', 'G4']/sum(tab[,'G4'])
T4sig <- tab['TRUE', 'T4']/sum(tab[,'T4'])

print(paste0('Percent significantly differentially expressed T4 v N4 is ', T4sig))
print(paste0('Percent significantly differentially expressed G4 v N4 is ', G4sig))

sum(tab[,'G4'])
sum(tab[,'T4'])

## This is so close the the normal numbers, we'll just report that.
length(which(bestT4$FDR < 0.05))
length(which(bestT4$FDR < 0.05))/nrow(bestT4)

length(which(bestG4$FDR < 0.05))
length(which(bestG4$FDR < 0.05))/nrow(bestG4)


```

So it looks like, particularly for T4, there is very little differential regulation. There may be some more with G4, but this is also a bit harder because G4 overlap N4 less with respect to CD69 expression.

Now, let's try enrichment testing.

```{r}
library(limma)
library(org.Mm.eg.db)
xx <- as.list(org.Mm.egENSEMBL2EG)

GO1 <- goana(resT4, species="Mm", geneid=xx[rownames(resT4)], FDR=0.05, trend=FALSE)
GO1tab <- topGO(GO1, ont="BP", number=Inf)
head(GO1tab)

GO2 <- goana(resG4, species="Mm", geneid=xx[rownames(resG4)], FDR=0.05, trend=FALSE)
GO2tab <- topGO(GO2, ont="BP", number=Inf)
head(GO2tab)

GOs <- list(T4=GO1tab, G4=GO2tab)
GOs <- lapply(GOs, function(x) {
  ps_list <- c(x$P.Up, x$P.Down)
  ps_adj <- p.adjust(ps_list, method="fdr")
  ps <- data.frame(ps_list, ps_adj)
  ps <- ps[order(ps$ps_list),]
  adjp <- max(which(ps$ps_adj < 0.05))
  adjp <- ps$ps_list[adjp]
  #adjp <- 0.05/nrow(x)
  return(x[(x$P.Up < adjp) | (x$P.Down < adjp),])
})

write.table(GOs[[1]], '../data/DE_combrunAT_spikenorm_filt05_T4vN4_GOsig_controlCD69sp2.txt', quote=FALSE, sep='\t')
write.table(GOs[[2]], '../data/DE_combrunAT_spikenorm_filt05_G4vN4_GOsig_controlCD69sp2.txt', quote=FALSE, sep='\t')

for(i in 1:length(GOs)){
  enrich <- apply(GOs[[i]], 1, function(x){
   if(as.numeric(x[6]) < as.numeric(x[7])){
      enrich <- -log10(as.numeric(x[6]))
    } else {
      enrich <- log10(as.numeric(x[7]))
    }
    return(enrich)   
    })
  GOs[[i]]$enrich <- enrich
}


## and plot out the top 20 for each

GOs20 <- lapply(GOs, function(x) {
  if(nrow(x)<20){
    return(x[order(x$enrich),])
  } else {
  top20 <- abs(x$enrich)[order(abs(x$enrich), decreasing=TRUE)][20]
  temp <- x[which(abs(x$enrich) >= top20),]
  return(temp[order(abs(temp$enrich)),])
  }
})

par(mar=c(5.1,22.1,2.1,2.1), mgp=c(2,1,0))
barplot(GOs20[[1]]$enrich, horiz=TRUE, beside=TRUE, main='T4 v N4 enrichment',
        col=c('green3'), las=1, xlab=expression(paste("-log"[10], " p-value x DE direction")),
        names.arg=GOs20[[1]]$Term, cex.names=0.8)

pdf('../data/DE_combrunAT_spikenorm_filt05_T4vN4_GO20_controlCD69sp2.pdf', height=3, width=8)
par(mar=c(4.1,22.1,2.1,2.1), mgp=c(2,1,0))
barplot(GOs20[[1]]$enrich, horiz=TRUE, beside=TRUE, main='',
        col=c('green3'), las=1, xlab='',
        names.arg=GOs20[[1]]$Term, cex.names=0.8)
dev.off()

par(mar=c(5.1,22.1,2.1,2.1), mgp=c(2,1,0))
barplot(GOs20[[2]]$enrich, horiz=TRUE, beside=TRUE, main='G4 v N4 enrichment',
        col=c('blue'), las=1, xlab=expression(paste("-log"[10], " p-value x DE direction")),
        names.arg=GOs20[[2]]$Term, cex.names=0.8)

pdf('../data/DE_combrunAT_spikenorm_filt10_G4vN4_GO20_controlCD69sp2.pdf', height=5, width=8)
par(mar=c(4.1,22.1,2.1,2.1), mgp=c(2,1,0))
barplot(GOs20[[2]]$enrich, horiz=TRUE, beside=TRUE, main='',
        col=c('blue'), las=1, xlab='',
        names.arg=GOs20[[2]]$Term, cex.names=0.8)
dev.off()

```

This has revealed some interesting genes.  Let's make some plots of shared genes, antigen presenting machinery (T4 enriched upregulated) and defense response (G4 enriched downregulated).

```{r}

plot_resid <- function(sceset, spline_mat, gene_symbol, 
                       conditions=c('N4_3h', 'N4_6h', 'T4_6h', 'G4_6h'), 
                       mylevels=c('N4_3h', 'N4_6h', 'T4_6h', 'G4_6h'),
                       mylevels2=c('N4', 'T4', 'G4'),
                       cols=c('red', 'green3', 'blue')){
  fdat <- rowData(sceset)
  rownames(fdat) <- rownames(sceset)
  genen <- rownames(fdat)[fdat$symbol %in% gene_symbol]
  cellinds <- which(colData(sceset)$Condition %in% conditions)
  df <- data.frame(Y=exprs(sceset)[genen,cellinds],
                   Cond=factor(colData(sceset[,cellinds])$Condition, levels=mylevels),
                   Batch=factor(colData(sceset[,cellinds])$batch),
                   Date=factor(colData(sceset[,cellinds])$Date))
  df$Cond2 <- unlist(lapply(strsplit(as.character(df$Cond), '_'), function(x){x[1]}))
  spline_mat <- spline_mat[cellinds,]
  resid <- lm(df$Y ~ df$Batch + df$Date:spline_mat)$residuals
  facs <- split(resid, factor(df$Cond2, levels=mylevels2))
  stripchart(facs, method='jitter', vertical=TRUE, pch=20, col=cols, las=1, 
             xlim=c(0.5,length(mylevels2)+0.5), 
             ylab=bquote('residual log'[2] ~ .(gene_symbol) ~ 'expression'))
  boxplot(facs, outline=FALSE, add=T, col=rgb(0,0,0, alpha=0), axes=F)
}

pdf('../data/DE_CombrunAT_spikenorm_filt05_CD69sp2DEsharedantiviral_genes.pdf', height=3, width=9)
par(mfrow=c(1,4), mgp=c(2,0.7,0), mar=c(3,3.5,1,0.2), bty='l', cex.lab=1.4, cex.axis=1.3)
plot_resid(sceAT, spline_mat=CD69spdf, 'Psmb8')
plot_resid(sceAT, spline_mat=CD69spdf, 'Tapbp')
plot_resid(sceAT, spline_mat=CD69spdf, 'Samhd1')
plot_resid(sceAT, spline_mat=CD69spdf, 'Ifi47')
dev.off()

pdf('../data/DE_CombrunAT_spikenorm_filt05_CD69sp2DET4classI_genes.pdf', height=3, width=9)
par(mfrow=c(1,4), mgp=c(2,0.7,0), mar=c(3,3.5,1,0.2), bty='l', cex.lab=1.4, cex.axis=1.3)
plot_resid(sceAT, spline_mat=CD69spdf, 'Tapbp')
plot_resid(sceAT, spline_mat=CD69spdf, 'Psmb8')
plot_resid(sceAT, spline_mat=CD69spdf, 'B2m')
plot_resid(sceAT, spline_mat=CD69spdf, 'H2-T23')
dev.off()

pdf('../data/DE_CombrunAT_spikenorm_filt05_CD69sp2DEG4chemokine_genes.pdf', height=3, width=9)
par(mfrow=c(1,4), mgp=c(2,0.7,0), mar=c(3,3.5,1,0.2), bty='l', cex.lab=1.4, cex.axis=1.3)
plot_resid(sceAT, spline_mat=CD69spdf, 'Xcl1')
plot_resid(sceAT, spline_mat=CD69spdf, 'Ccl3')
plot_resid(sceAT, spline_mat=CD69spdf, 'Ccl4')
plot_resid(sceAT, spline_mat=CD69spdf, 'Ifng')
dev.off()

pdf('../data/DE_CombrunAT_spikenorm_filt05_CD69sp2_GZMB.pdf', height=4, width=2.7)
par(mgp=c(2,0.7,0), mar=c(3,3.5,1,0.2), bty='l', cex.lab=1.4, cex.axis=1.3, cex=0.9)
plot_resid(sceAT, spline_mat=CD69spdf, 'Gzmb')
dev.off()

```

And finally we want to check how these genes act in general. Are they upregulated with activation? Downregulated? Unchanged?

```{r}
expr_box_plot <- function(gene_symbol, sceset, 
                     split_facs=c('unstimulated', 'N4_3h', 'N4_6h', 'T4_6h', 'G4_6h', 'NP68_6h'), 
                     name_facs=c('unstimulated', 'N4_3h', 'N4_6h', 'T4_6h', 'G4_6h', 'NP68_6h'), 
                     cols=c('grey', 'orange', 'red', 'green3', 'blue', 'grey40')){
  fdat <- rowData(sceset)
  rownames(fdat) <- rownames(sceset)
  genen <- rownames(fdat)[fdat$symbol %in% gene_symbol]
  temp <- split(exprs(sceset)[genen,], 
                factor(as.character(colData(sceset)$Condition), levels=split_facs))
  names(temp) <- name_facs
  stripchart(temp, method='jitter', vertical=TRUE, pch=20, col=cols, las=2, 
             xlim=c(0.5,length(temp)+0.5), 
             ylab=bquote('log'[2] ~ .(gene_symbol) ~ 'expression'))
  boxplot(temp, outline=FALSE, add=T, col=rgb(0,0,0, alpha=0), axes=F)
  
}

## change sceset arguments below to sceAT???

par(mfrow=c(1,4), mgp=c(2,0.5,0), mar=c(6,3.5,1,0.2), bty='l', cex.lab=1.5, cex.axis=1)
expr_box_plot('Tapbp', sceset=sceAT)
expr_box_plot('Psmb8', sceset=sceAT)
expr_box_plot('B2m', sceset=sceAT)
expr_box_plot('H2-T23', sceset=sceAT)

par(mfrow=c(1,4), mgp=c(2,0.5,0), mar=c(3,3.5,1,0.2), bty='l', cex.lab=1.5, cex.axis=1)
expr_box_plot('Xcl1', sceset=sceAT)
expr_box_plot('Ccl3', sceset=sceAT)
expr_box_plot('Ccl4', sceset=sceAT)
expr_box_plot('Ifng', sceset=sceAT)

pdf('../data/DE_CombrunAT_spikenorm_filt05_CD69sp2DET4classI_genes_expr.pdf', height=3, width=10)
par(mfrow=c(1,4), mgp=c(1.2,0.5,0), mar=c(6,3.5,1,0.2), bty='l', cex.lab=1.2, cex.axis=1)
expr_box_plot('Tapbp', sceset=sceAT)
expr_box_plot('Psmb8', sceset=sceAT)
expr_box_plot('B2m', sceset=sceAT)
expr_box_plot('H2-T23', sceset=sceAT)
dev.off()

pdf('../data/DE_CombrunAT_spikenorm_filt05_CD69sp2DEG4chemokine_genes_expr.pdf', height=3, width=10)
par(mfrow=c(1,4), mgp=c(1.2,0.5,0), mar=c(6,3.5,1,0.2), bty='l', cex.lab=1.2, cex.axis=1)
expr_box_plot('Xcl1', sceset=sceAT)
expr_box_plot('Ccl3', sceset=sceAT)
expr_box_plot('Ccl4', sceset=sceAT)
expr_box_plot('Ifng', sceset=sceAT)
dev.off()

pdf('../data/DE_CombrunAT_spikenorm_filt05_CD69sp2_GZMB_expr.pdf', height=4, width=3.5)
par(mgp=c(1.2,0.5,0), mar=c(6,3.5,1,0.2), bty='l', cex.lab=1.2, cex.axis=1, cex=0.9)
expr_box_plot('Gzmb', sceset=sceAT)
dev.off()


```

So class I presentation genes are upregulated even beyond what is seen in N4 condition, while chemokines and cytokines just fail to turn on in G4 despite some level of activation. But in general, very little is differentially expressed given activation status.

```{r}
sessionInfo()
```