---
title: "CD8_APLcombined_unstimulated_noise"
author: "Arianne Richard"
date: "April 2018"
output: html_document
---

This file searches for structure or hints of what may make certain cells better responders in the unstimulated condition.

First we load the data, remove uninformative genes, and normalize the data.

```{r}
library(scater)
library(scran)
library(SingleCellExperiment)
library(RColorBrewer)
library(gplots)
library(dpt)
source('dpt_plots.R')
```


We'll look within the single batch where the unstimulated cells were and look at pseudotime from the original N4 timecourse. We'll first process it as we did previously.

```{r}
sce <- readRDS('../data/sce_run2.rds')
drop_filt <- rowSums(counts(sce) > 0) >= (ncol(sce) * 0.05)
sce <- sce[drop_filt,]
mean_filt <- rowMeans(counts(sce)) >= 1
sce <- sce[mean_filt,]

sceunind <- sce[,sce$Condition %in% 'unstimulated']
sceunind <- computeSpikeFactors(sceunind, general.use=TRUE)
sceunind <- normalize(sceunind, exprs_values='counts', return_norm_as_exprs=TRUE, log_exprs_offset=1)

sce <- computeSpikeFactors(sce, general.use=TRUE)
sce <- normalize(sce, exprs_values='counts', return_norm_as_exprs=TRUE, log_exprs_offset=1)

```

# Highly variable gene analysis

What are the highly variable genes?

```{r}

df <- data.frame(batch=as.character(sceunind$Sorting.Plate))
mod <- model.matrix(~0+batch, data=df)
alt.fit <- trendVar(sceunind, use.spikes=TRUE, design=mod, method='loess', parametric='true', span=0.95) 

# decompose with the spikes so we can plot it out
alt.decomp2 <- decomposeVar(sceunind, alt.fit, design=mod, get.spikes=TRUE)
plot(alt.decomp2$mean, alt.decomp2$total, xlab="Mean log-expression", ylab="Variance")
points(alt.decomp2$mean[which(isSpike(sceunind))], alt.decomp2$total[which(isSpike(sceunind))], col='red', pch=20)
o <- order(alt.decomp2$mean)
lines(alt.decomp2$mean[o], alt.decomp2$tech[o], col="red", lwd=2)

# and decompose without spikes to get the top highly variable genes with indices to use later
alt.decomp <- decomposeVar(sceunind, alt.fit, design=mod, get.spikes=FALSE)
fdat <- rowData(sceunind)
rownames(fdat) <- fdat$GeneID
alt.decomp$symbol <- fdat[rownames(alt.decomp), 'symbol']
top.hvgs <- order(alt.decomp$bio, decreasing=TRUE)
head(alt.decomp[top.hvgs,], 20)

```

Ordering cells by decreasing biological variability reveals:

1. Top gene Dusp1 is a Mapk phosphatase - this could be interesting with respect to signalling initiation.

2. Adhesion molecule Itgb2 is CD18, part of LFA-1, which is known to play a role in T cell activation via adhesion.

3. Cd6 is known to have a role in T cell costimulation

4. Jun is part of AP-1 transcription factors.

Let's look at the top 9 and particularly at LFA-1 components:

```{r}

expr_box_plot <- function(gene_symbol, sceset, 
                     split_facs=c('unstimulated', 'N4_1h', 'N4_3h', 'N4_6h'), 
                     name_facs=c('unstimulated', 'N4_1h', 'N4_3h', 'N4_6h'), 
                     cols=c('grey', 'gold', 'orange', 'red')){
  fdat <- rowData(sceset)
  rownames(fdat) <- fdat$GeneID
  genen <- rownames(fdat)[fdat$symbol %in% gene_symbol]
  temp <- split(exprs(sceset)[genen,], 
                factor(as.character(colData(sceset)$Condition), levels=split_facs))
  names(temp) <- name_facs
  stripchart(temp, method='jitter', vertical=TRUE, pch=20, col=cols, las=2, 
             xlim=c(0.5,length(temp)+0.5), 
             ylab=bquote('log'[2] ~ .(gene_symbol) ~ 'expression'))
  boxplot(temp, outline=FALSE, add=T, col=rgb(0,0,0, alpha=0), axes=F)
}

par(mfrow=c(3,3), mgp=c(2,0.5,0), mar=c(6,3.5,1,0.2), bty='l', cex.lab=1.5, cex.axis=1)
for (i in c(1:9)){
  expr_box_plot(as.character(alt.decomp[top.hvgs,'symbol'][i]), sceset=sce)
}

par(mfrow=c(1,2), mgp=c(2,0.5,0), mar=c(6,3.5,1,0.2), bty='l', cex.lab=1.5, cex.axis=1)
expr_box_plot('Itgal', sceset=sce)
expr_box_plot('Itgb2', sceset=sce)

```

# Differential expression along pseudotime among the unstimulated cells

We first need to batch correct the whole timecourse dataset as we did previously.

```{r}

library(sva)
mod <- model.matrix(~factor(as.character(sce$Condition)))
combat_sce <- ComBat(dat=exprs(sce), batch=colData(sce)$Sorting.Plate,
                     mod=mod, par.prior=TRUE, prior.plots=FALSE)
exprs(sce) <- combat_sce
sce <- sce[, sce$Condition %in% c('unstimulated', 'N4_1h', 'N4_3h', 'N4_6h')]

```

Now we can look at pseudotime.

```{r}
## remove spikes as these are no longer needed
sce <- sce[!isSpike(sce),]

library(destiny)

resTC <- quick_dpt(sce, conditions=c('unstimulated', 'N4_1h', 'N4_3h', 'N4_6h'), col=c('grey', 'gold', 'orange', 'red'), base_condition='unstimulated', stim_condition='N4_6h')

resTC[[1]]

```

What is differentially expressed along pseudotime (considered as a continuous variable) among the unstimulated cells?

```{r}
library(edgeR)

## redefine unstimulated indices based on new, limited sceset
unind <- which(sce$Condition %in% 'unstimulated')
sceunind <- sce[,unind]

df <- data.frame(batch=as.character(sceunind$Sorting.Plate), 
                 dpt=resTC$pt$DPT[unind])
mod <- model.matrix(~0+dpt+batch, data=df)
z_s <- convertTo(sceunind, type="edgeR")
y_s <- estimateDisp(z_s, design=mod)
fit <- glmFit(z_s, design=mod, dispersion=y_s$tagwise.dispersion, prior.count=1)
res <- glmLRT(fit, coef='dpt')
bestun <- data.frame(topTags(res, n=nrow(z_s)))
bestun$symbol <- as.character(fdat[rownames(bestun),'symbol'])
head(bestun, 20)

plot(resTC$pt$DPT[unind], 
     exprs(sceunind)[rownames(bestun)[1], ], 
     ylab=bestun[1, 'symbol'],
     xlab='dpt')

```

The only significant association is with Ift52. However, while it is associated with some of those advanced cells, it is certainly not deterministic in defining cells that are more progressive along pseudotime.

# tSNE analysis

Let's look at whether there is any structure by tSNE in the unstimulated cells.

```{r}

library(Rtsne)
tn <- Rtsne(t(exprs(sceunind)), perplexity=5)
plot(tn$Y[,1], tn$Y[,2])

```

There is basically no structure here either.


# Clustering

And let's hierarchically cluster the cells to see if any structure is apparent.

```{r}
sceunind_var <- sceunind[top.hvgs[1:100],]
## note - these genes were calculated using the dataset with batch as an uninteresting variance-contributing factor but before any batch correction and with normalization factors centered on the unstimulated cells only; we are now taking the unstimulated cells from the full normalized and batch-corrected dataset and filtering for these genes
library(pheatmap)

col_anno <- data.frame(pseudotime=resTC$pt$DPT[unind], batch=sceunind_var$Sorting.Plate)
rownames(col_anno) <- colnames(sceunind_var)

hmcol <- colorRampPalette(c('blue', 'white', 'red'))(200)

pheatmap(exprs(sceunind_var), 
         annotation_col=col_anno, 
         show_rownames=FALSE, 
         show_colnames=FALSE, 
         color=hmcol)

pdf('../data/APL_TC_Unstim_HVG_heaptmap2.pdf', height=5, width=5)
pheatmap(exprs(sceunind_var), 
         annotation_col=col_anno, 
         show_rownames=FALSE, 
         show_colnames=FALSE, 
         color=hmcol)
dev.off()
```

```{r}
sessionInfo()
```