---
title: "APL1_cydar"
author: "Arianne Richard"
date: "May 2018"
output: html_document
---

This script should run all necessary preprocessing functions and differential abundance analysis on CD8+ T cell stimulation CyTOF data available in the community version of cytobank, accession number 66456.

# Load the data

First we load in the data set and appropriately name all of the channels and samples.

```{r}
library(ncdfFlow)
library(cydar)
library(flowCore)
f <- list.files(path='data', pattern='*0.fcs', full.names=TRUE)
ncdf <- read.ncdfFlowSet(files=f, ncdfFile='ncdf')

channels <- data.frame(channel=c("Yb174Di", "Gd156Di", "Nd150Di", "Tm169Di", "Er170Di", "Tb159Di", "Er168Di", "Gd155Di", "Eu151Di", "Y89Di", "Nd144Di", "Dy163Di", "Sm154Di", "Yb172Di", "Yb173Di", "Gd158Di", "Dy161Di", "Ho165Di", "Gd160Di"), 
                       marker=c("174Yb_IL-7Ra", "156Gd_BTLA", "150Nd_CD27", "169Tm_TCRb", "170Er_NK1.1", "159Tb_PD-1", "168Er_CD8a", "155Gd_CD44", "151Eu_CD25", "89Y_CD45", "144Nd_CD11a-FITC", "163Dy_CD62L-APC", "154Sm_CTLA-4", "172Yb_Perforin", "173Yb_Granzyme B", "158Gd_Foxp3", "161Dy_Tbet", "165Ho_IFNg", "160Gd_LAMP-1-PE"))

samples <- read.table('data/experiment_66456_annotations.tsv', sep='\t', header=TRUE)
rownames(samples) <- samples$FCS.Filename

pdat <- pData(ncdf)
pdat$treatment <- samples[rownames(pdat),'Conditions']
pdat$individual <- samples[rownames(pdat),'Individuals']
pData(ncdf) <- pdat

```

# Setting up the CD8+ T cell gating strategy on all cells pooled

Now we want to normalize the data and remove clogs, debris, dead cells, etc.
First pool all of the cells and deal with the biggest machine clog.

```{r}
pool.ff <- poolCells(ncdf, equalize=FALSE)
plot(data.frame(exprs(pool.ff))$Time, log10(data.frame(exprs(pool.ff))$Y89Di), cex=0.1, xlab='time', ylab='CD45', ylim=c(-1,4))
abline(v=2700000, col='dodgerblue')
abline(v=2950000, col='dodgerblue')
```

Remove cells within this time gate and re-plot to make sure we removed the right thing.

```{r}
clog <- rectangleGate(filterId="timegate", list("Time"=c(2700000,2950000)))
clogfilt <- filter(pool.ff, clog)
pool.ff <- split(pool.ff, clogfilt)['timegate-'][[1]]

plot(data.frame(exprs(pool.ff))$Time, log10(data.frame(exprs(pool.ff))$Y89Di), cex=0.1, xlab='time', ylab='CD45', ylim=c(-1,4))
```

Now let's transform the data and find the beads.

```{r}
pool.ff <- pool.ff[,!colnames(pool.ff) %in% c('Event_length', 'Time', 'Offset', 'Center', 'Width', 'Residual')]

lgcl <- logicleTransform( w = 0.1, t= 10000, m =4.5) ## using default except lowered w
trans <- transformList(colnames(pool.ff), lgcl)
proc.ff <- transform(pool.ff, trans)
plot(data.frame(exprs(proc.ff))$Eu151Di, data.frame(exprs(proc.ff))$Eu153Di, xlab='Eu151', ylab='Eu153')
```

Remove these beads and again re-plot.

```{r}
beads <- rectangleGate(filterId="beadgate", list("Eu153Di"=c(2.5, Inf),
"Eu153Di"=c(2.5, Inf)))

beadfilt <- filter(proc.ff, beads)
filt_set <- split(proc.ff, beadfilt)['beadgate-'][[1]]
plot(data.frame(exprs(filt_set))$Eu151Di, data.frame(exprs(filt_set))$Eu153Di, xlab='Eu151', ylab='Eu153')
```

Gate on Ir DNA markers for singlets.  Unfortunately, most of this data is in doublets so we'll use some options from dnaGate to select the smaller peak.

```{r}
hist(data.frame(exprs(filt_set))$Ir191Di, breaks=200, xlab='Ir191', main='')
hist(data.frame(exprs(filt_set))$Ir193Di, breaks=200, xlab='Ir193', main='')
dna <- dnaGate(filt_set, 'Ir191Di', 'Ir193Di', type='both', rank=2)
smoothScatter(data.frame(exprs(filt_set))$Ir191Di, data.frame(exprs(filt_set))$Ir193Di)
polygon(dna@boundaries[,1], dna@boundaries[,2], border="red")

```

Apply these gates in both channels

```{r}
dnafilt <- filter(filt_set, dna)
filt_set <- Subset(filt_set, dnafilt)
hist(data.frame(exprs(filt_set))$Ir191Di, breaks=200, xlab='Ir191', main='')
hist(data.frame(exprs(filt_set))$Ir193Di, breaks=200, xlab='Ir193', main='')
```

Exclude dead cells as Pt high. We'll do this manually because we want to be very stringent.

```{r}
hist(data.frame(exprs(filt_set))$Pt195Di, breaks=200, xlab='Pt195', main='')
abline(v=2, col='dodgerblue')
```

Retain only cells that are Pt195Di low.

```{r}
live <- rectangleGate(filterId="livegate", list("Pt195Di"=c(0, 2)))
livefilt <- filter(filt_set, live)
filt_set <- Subset(filt_set, livefilt)
hist(data.frame(exprs(filt_set))$Pt195Di, breaks=200, xlab='Pt195', main='')
```

Exclude extreme low values of CD45 as non-haematopoietic.

```{r}
hist(data.frame(exprs(filt_set))$Y89Di, breaks=200, xlab='CD45', main='')
```

Gate out 5 MAD lower - just removing the properly negative ones.

```{r}
haem <- outlierGate(filt_set, 'Y89Di', nmads=5, type='lower')
haemfilt <- filter(filt_set, haem)
filt_set <- Subset(filt_set, haemfilt)
hist(data.frame(exprs(filt_set))$Y89Di, breaks=200, xlab='CD45', main='')
```

Remove non-T cells.

```{r}
hist(data.frame(exprs(filt_set))$Tm169Di, breaks=200, xlab='TCRb', main='')
```

Gate out 6 MAD lower because TCR levels can fall dramatically in activating cells and we don't want to miss these. (Also, we started with primarily CD8+ T cells in the live fraction by experimental design.)

```{r}
Tcell <- outlierGate(filt_set, 'Tm169Di', nmads=6, type='lower')
Tfilt <- filter(filt_set, Tcell)
filt_set <- Subset(filt_set, Tfilt)
hist(data.frame(exprs(filt_set))$Tm169Di, breaks=200, xlab='TCRb', main='')
```

Remove non-CD8+ T cells.

```{r}
hist(data.frame(exprs(filt_set))$Er168Di, breaks=200, xlab='CD8a', main='')
```

Gate out 5 MAD lower because this can also drop dramatically with activation. (Also, we started with primarily CD8+ T cells in the live fraction by experimental design.)

```{r}
CD8 <- outlierGate(filt_set, 'Er168Di', nmads=5, type='lower')
CD8filt <- filter(filt_set, CD8)
filt_set <- Subset(filt_set, CD8filt)
hist(data.frame(exprs(filt_set))$Er168Di, breaks=200, xlab='CD8a', main='')
```

# Applying the gating strategy to individual samples

Now apply these same gates to the separated samples.
First the clog filter and visualize just sample 2 to see that the filter worked.

```{r}
processed.exprs <- ncdf
clogfilt <- filter(processed.exprs, clog)
processed.exprs <- split(processed.exprs, clogfilt)['timegate-'][[1]]
plot(data.frame(exprs(processed.exprs[[2]]))$Time, log10(data.frame(exprs(processed.exprs[[2]]))$Y89Di), cex=0.1, xlab='time', ylab='CD45', ylim=c(-1,4))
```

Now apply the transformation and apply the bead gate (again visualizing sample 2).

```{r}
processed.exprs <- processed.exprs[,!colnames(processed.exprs) %in% c('Event_length', 'Time', 'Offset', 'Center', 'Width', 'Residual')]
processed.exprs <- transform(processed.exprs, trans)

beadfilt <- filter(processed.exprs, beads)
processed.exprs <- split(processed.exprs, beadfilt)['beadgate-'][[1]]
plot(data.frame(exprs(processed.exprs[[2]]))$Eu151Di, data.frame(exprs(processed.exprs[[2]]))$Eu153Di, xlab='Eu151', ylab='Eu153')
```

Now apply the DNA gate (again visualizing sample 2).

```{r}
processed.exprs <- Subset(processed.exprs, dna)
hist(data.frame(exprs(processed.exprs[[2]]))$Ir191Di, breaks=100, xlab='Ir191', main='')
hist(data.frame(exprs(processed.exprs[[2]]))$Ir193Di, breaks=100, xlab='Ir193', main='')
```

And the live cell gate (again visualizing sample 2).

```{r}
processed.exprs <- Subset(processed.exprs, live)
hist(data.frame(exprs(processed.exprs[[2]]))$Pt195Di, breaks=100, xlab='Pt195', main='')
```

And the CD45 haematopoietic cell gate.

```{r}
processed.exprs <- Subset(processed.exprs, haem)
hist(data.frame(exprs(processed.exprs[[2]]))$Y89Di, breaks=100, xlab='CD45', main='')
```

And the TCRb T cell gate.

```{r}
processed.exprs <- Subset(processed.exprs, Tcell)
hist(data.frame(exprs(processed.exprs[[2]]))$Tm169Di, breaks=100, xlab='TCRb', main='')
```

And the CD8+ cell gate.

```{r}
processed.exprs <- Subset(processed.exprs, CD8)
hist(data.frame(exprs(processed.exprs[[2]]))$Er168Di, breaks=100, xlab='CD8a', main='')
```

## And perform differential abundance analysis

Now we can begin running CyDAR analysis. First, limit the markers to those used for proteins. (We're not going to exclude the markers we filtered on because changes in expression of CD45, TCRb and CD8 correspond to activation status and can be interesting! But we want to get rid of DNA, live/dead, etc). Then we look at neighbor distances to determine if our tolerance default of 0.5 is okay.

```{r}
processed.exprs <- processed.exprs[,colnames(processed.exprs) %in% channels[,1]]
cd <- prepareCellData(processed.exprs, naive=FALSE)
dist <- neighborDistances(cd)
boxplot(dist, xlab="Number of neighbors", ylab="Tolerance")
abline(h=0.5, col="red", lwd=2, lty=2)
```

This looks fine. We have even more cells per hypersphere than needed. Now count cells into hyperspheres, considering only hyperspheres with 5 or more cells, and run statistical analysis under the null hypothesis that none of the ovalbumin-derived peptides has an effect.

```{r}
cd <- countCells(cd, tol=0.5)
peptide <- samples[colnames(cd), 'Conditions']
peptide <- unlist(lapply(strsplit(as.character(peptide), ' '), function(x) {x[1]}))
mouse <- samples[colnames(cd), 'Individuals']
mouse <- gsub(' ', '', mouse)
design <- model.matrix(~ 0 + peptide + mouse)

library(edgeR)
y <- DGEList(assay(cd), lib.size=cd$totals)
keep <- aveLogCPM(y) >= aveLogCPM(5, mean(cd$totals))
table(keep)
cd2 <- cd[keep,]
y <- y[keep,]

adjc <- edgeR::cpm(y, log=TRUE, prior.count=3)
label <- as.character(samples$Conditions)
col <- c("blue", "red", "black", "cyan2", "green3")[factor(peptide)]
plotMDS(adjc, label=label, col=col)

y <- estimateDisp(y, design)
plotBCV(y)

fit <- glmQLFit(y, design, robust=TRUE)
plotQLDisp(fit)

con <- makeContrasts(N4=peptideN4 - peptideNP68,
                     G4=peptideG4 - peptideNP68,
                     Q4H7=peptideQ4H7 - peptideNP68,
                     T4=peptideT4 - peptideNP68, levels=design)

res.all <- glmQLFTest(fit, contrast=con)
qval.all <- spatialFDR(intensities(cd2), res.all$table$PValue)
is.sig.all <- qval.all <= 0.05
summary(is.sig.all)

```

And let's save this data.

```{r}
save(list=ls(), file='data/cytof_analysis_in_progress.RData')

sessionInfo()
```

